---
title: "Data analysis challenge"
author: "Peter Thorpe"
date: "26th Oct 2022"
output:
  pdf_document: default
  html_document: default

---

```{r include=FALSE}

knitr::opts_chunk$set(comment = NA)
error = TRUE

```


```{r, global_options, include=FALSE}
knitr::opts_chunk$set(message=FALSE, tidy.opts=list(width.cutoff=80), tidy=TRUE) 

```
## 1 the problem

Data recived however, missing the coloumn names - data from Antartica. 

## 2 time series data 


105032 obervations recorded in 5 different ways ... 
lets try and understand the properties: problems, trends corrolation ... and so on. 

***

## Load the library needed

```{r}

library(knitr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(readr)
library(magrittr)
library(devtools)
# install_github("vqv/ggbiplot")
library(ggbiplot)
library(Hmisc)
library(corrplot)



```


***

##  Load the data

counts were already generated using salmon and counts.matrix generated using trinity. 

```{r}
setwd("C:/Users/pjt6/Documents/DAG_challenge")

# check it
getwd()

```
***


The counts data in contained in the counts.matrix, each gene has a digital count per
condition/ rep

```{r}

# load in the data
time_series_data <- read.table("data/data.csv", sep=",", header = TRUE, row.names=NULL)


summary(time_series_data)

```
\newpage

have a quick look at the data:

```{r}
head(time_series_data)

```

```{r}

# Using na.omit and assigning to another data frame
time_series_data_without_NAs <- na.omit(time_series_data)

```


check for NA in the dataset.

```{r}

# Checking the dimensions with the dim() function (rows then columns)
dim(time_series_data)

dim(time_series_data_without_NAs)

```


There appears to be no NAs in the data. 
now check for duplicated rows. 


```{r}

# duplicated(time_series_data)  # dont run this!
sum (duplicated(time_series_data))


```


there are no duplicated rows. 


```{r}
# this doesnt help
boxplot(time_series_data$A)



```


lets do a PCA to see how the variables separate out. 

```{r}

time_series_data.pca <- prcomp(time_series_data[,c(1,2,3,4,5)], center = TRUE,scale. = TRUE)


# ggbiplot(time_series_data.pca)
# this wont install and run on my windows laptop!

#############################################
# library/3.6/gtable/R/gtable.rdb' is corrupt

```


lets have a look at some corrolation matrix between these 

```{r}
# get the first 5 cols. 
time_series_data_w_last_clo <- time_series_data[,c(1,2,3,4,5)]

# get a corrolation matrix with P values
time_series.rcorr = rcorr(as.matrix(time_series_data_w_last_clo))

# see the results
time_series.rcorr



```


none of the corrolations are significant, see plot below. However, there is some loose corrolation. 


```{r}
	
time_series.cor = cor(time_series_data_w_last_clo, method = c("spearman"))
	
corrplot(as.matrix(time_series.cor))

```

look at a quick plot A vs ID

```{r}

ggplot(data = time_series_data, aes(x = A, y = id)) +
  geom_line(linetype = "dashed")+
  geom_point()




```


Ok, so this is not a nice plot. Lets randomly subsample this to see if we can see any better trends. 


```{r}

time_series_data_subsample <- time_series_data[sample(nrow(time_series_data), 1000), ]

ggplot(data = time_series_data_subsample, aes(x = A, y = id)) +
  geom_line(linetype = "dashed")+
  geom_point()

```

I dont know if there may be multiple peaks in this data set. So I will try a Fourier Transform.
I know from my NMR days that I should be able to see peaks if there are any in the data here,
but my code is not working. 


```{r}

# apply(time_series_data, 2,function(x) fft(as.numeric(x)))

#A_fft <- as.data.frame(fft(as.numeric(time_series_data$A)))

#A_fft <- unlist(A_fft)


#ggplot(data = A_fft, aes(x = A_fft, y = time_series_data$id)) +
#  geom_line(linetype = "dashed")+
#  geom_point()

```



## 3 LOOK AT COL B 


look at a quick plot B vs ID

```{r}

ggplot(data = time_series_data, aes(x = B, y = id)) +
  geom_line(linetype = "dashed")+
  geom_point()



```


That doesn thelp - lets try a histogram


```{r}


ggplot(data = time_series_data, aes(x=B)) + 
  geom_histogram(binwidth=1)


```

Ok so we have a poisson distribution. These could be counts of observations per time point. 


```{r}


# Extract histogram information
hist_vec <- hist(time_series_data$B)   
  
# Store histogram counts in frequency
frequency <- hist_vec$counts

frequency


```


## 3 LOOK AT COL C 


look at a quick plot B vs ID

```{r}

ggplot(data = time_series_data, aes(x = C, y = id)) +
  geom_line(linetype = "dashed")+
  geom_point()



```

Ok, lets try a histogram of this. 

```{r}


# Extract histogram information
hist_vec <- hist(time_series_data$C)   
  
# Store histogram counts in frequency
frequency <- hist_vec$counts

frequency


```




 Lets randomly subsample this to see if we can see any better trends. (the number to subsample by could be better chosen)


```{r}

time_series_data_subsample <- time_series_data[sample(nrow(time_series_data), 1000), ]

ggplot(data = time_series_data_subsample, aes(x = C, y = id)) +
  geom_line(linetype = "dashed")+
  geom_point()

```





## 3 LOOK AT COL D 


look at a quick plot D vs ID

```{r}

ggplot(data = time_series_data, aes(x = D, y = id)) +
  geom_line(linetype = "dashed")+
  geom_point()



```


 Lets randomly subsample this to see if we can see any better trends. (the number to subsample by could be better chosen)


```{r}

time_series_data_subsample <- time_series_data[sample(nrow(time_series_data), 1000), ]

ggplot(data = time_series_data_subsample, aes(x = D, y = id)) +
  geom_line(linetype = "dashed")+
  geom_point()

```



```{r}
# this doesnt help
boxplot(time_series_data$D)



```



Ok, lets try a histogram of this. 

```{r}


# Extract histogram information
hist_vec <- hist(time_series_data$D)   
  
# Store histogram counts in frequency
frequency <- hist_vec$counts

frequency

```


## SUMMARY colA

My hyposthesis for ColA is that is it temperature data over a year, based on the shape of the graph (although this graph is horrible - changing the defaults would be better) and based on the values. Min -7, mean of 9. The shape of the graph suggests to me the change in temp over the year. Given that there are 3.154e+7 seconds in a year and 105032 obseravtions. 

print(3.154e+7 / 105032) = 300

300/60 =  5.

Therefore there is an observation every 5 mins. 



QUESTION TO THE REASEARCH TEAM/ problems: These temperatures (if it is temp data) seem a little high for Antartica. Is the equipment callibrated correctly? I know there are serious warming problems, how does this compare to last year?



## SUMMARY colB

This is a poisson distribution. These are discreet counts of observations per time point. ColA and ColB have a 0.11 corrolation, thus they are not corrolated.  


problem: Looking at the plot there is a periodical increase then sharpe decrease. The pattern looks too regular to be biological data? Systematic error?

## SUMMARY colC


```{r}
summary(time_series_data)

```

ColA and C have a corroltation of 0.46. This is somewhat corrolated, although not statistically significant. 

QUESTION TO THE REASEARCH TEAM/ Problems: I am concerned that an element from ColD has ended up in ColC (max: 1098) .  Looking at the medians of both cols and the max in ColC. Although, there are a few other high enteries in ColC which may be out of place in ColD ... This needs to be clarified. 


## SUMMARY colD
There is a slight negative screw to the data with several very low values. 



