---
title: "Data analysis challenge"
author: "Peter Thorpe"
date: "26th Oct 2022"
output:
  pdf_document: default
  html_document: default

---

```{r include=FALSE}

knitr::opts_chunk$set(comment = NA)
error = TRUE

```


```{r, global_options, include=FALSE}
knitr::opts_chunk$set(message=FALSE, tidy.opts=list(width.cutoff=80), tidy=TRUE) 

```
## 0.1 time series data 


105032 obervations recorded in 5 different ways ... 
lets try and understand the properties: problems, trends corrolation ... and so on. 

***

## Load the library needed

```{r}

library(knitr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(readr)
library(magrittr)
library(devtools)
# install_github("vqv/ggbiplot")
library(ggbiplot)
library(Hmisc)
library(corrplot)



```


***

##  Load the data

counts were already generated using salmon and counts.matrix generated using trinity. 

```{r}
setwd("C:/Users/pjt6/Desktop/data_task")

# check it
getwd()

```
***


The counts data in contained in the counts.matrix, each gene has a digital count per
condition/ rep

```{r}

# load in the data
time_series_data <- read.table("data/data.csv", sep=",", header = TRUE, row.names=NULL)


summary(time_series_data)

```
\newpage

have a quick look at the data:

```{r}
head(time_series_data)

```

```{r}

# Using na.omit and assigning to another data frame
time_series_data_without_NAs <- na.omit(time_series_data)

```


check for NA in the dataset.

```{r}

# Checking the dimensions with the dim() function (rows then columns)
dim(time_series_data)

dim(time_series_data_without_NAs)

```


There appears to be no NAs in the data. 
now check for duplicated rows. 


```{r}

duplicated(time_series_data)
sum (duplicated(time_series_data))


```


there are no duplicated rows. 


```{r}
# this doesnt help
boxplot(time_series_data$A)
